\documentclass{tufte-handout}

% Be sure to check and update this:
\def\PuTTYurl{https://the.earth.li/~sgtatham/putty/0.74/w32/putty-0.74-installer.msi}
\def\GlobalProtectUrl{https://kb.northwestern.edu/page.php?id=94726}
\def\PuTTYhelpUrl{https://superuser.com/questions/1159811/openssh-proxycommand-equivalent-in-putty}
\def\setupScriptUrl{https://bit.ly/2RXwrkb}
\def\setupScriptLink{\href{\setupScriptUrl}{\setupScriptUrl}}

\usepackage{211lab}

\LabInfo{1}{Navigating the Unix Shell}

\begin{document}

\maketitle

Today we are going over the basics of how to log into a remote computer,
use shell commands to create and edit files, and compile and run C code.
\marginnote{There are different shells with slightly different syntax.
In this class, we will use a shell called \progname{fish} (for
``Friendly Interactive SHell''). Other shells include \progname{tcsh}
(the default on the EECS servers), \progname{bash} (the default on most
Linux systems), and \progname{zsh} (the default on recent versions of
Mac OS).}

The shell works as a textual conversation. It presents a prompt, like
\verb![wsc147@robin]~/cs211%!. (The default EECS prompt shows the
username, the hostname, and the current working directory.) You type a
command and press enter. The shell executes the command and then prints
another prompt, waiting for further commands. For example, to list the
files in the current directory, you will run the \progname{ls} command
by typing it at the prompt:

\begin{CmdLine*}
  \C ls \\
\end{CmdLine*}

\marginnote[-\baselineskip]{Don't type the \prompt. That represents the
prompt that the shell prints for you to tell you it's ready.}

\noindent
Before you can do that, though, we have to get you logged in.

\section{Getting around the firewall}

The workstations we want to connect to are not directly accessible from
off-campus, so we will need to do a little work to reach them. There are
essentially two alternatives:

\begin{itemize}

  \item Use Northwestern’s virtual private network (VPN). When you
    enable this, all your network traffic goes through an encrypted
    tunnel to Northwestern’s campus, which means you are effectively
    there. This solution is easy, but it can slow down your network, so
    you’ll want to turn it off when you aren’t using it.

    To setup the NU VPN, follow \href{\GlobalProtectUrl}{these
    instructions}.

  \item Configure SSH to ``proxy jump'' through a different host than
    the one we ultimately want. SSH can do this automatically for you if
    you set it up just right, but it’s a bit harder than using the VPN.
    Instructions for setting up a proxy appear in the per-system
    sections below.

\end{itemize}

\section{Logging in}

For the majority of you who are unfamiliar with the Unix shell, it
probably seems like a scary foreign concept reserved for computer
hackers on TV shows and movies.  However, in reality, with a little bit
of time and a few basic commands, you will realize that the Unix shell
is not something to be scared of, and in fact a very useful tool to
embrace as you continue your computer science education. Don't get
frustrated if it seems hard at first! Every great computer scientist was
at one point also unfamiliar with the shell, just like you, but with a
little bit of exposure, it will start to make sense.

SSH (secure shell) is a protocol that allows you to login remotely onto
an external system. We will be using it in order to create a connection
onto a Northwestern remote server, where we will be learning our first
Unix skills. For the first step of establishing the connection, it will
be different for Windows and Mac/Linux, but for the rest it should not
matter which OS you are on, since you'll be using the remote Unix
machine.

\subsection{Windows}

Download the \href{\PuTTYurl}{SSH client PuTTY}%
\marginnote{\url{\PuTTYurl}}; we recommend the
MSI installer, since it’s usually easier.
The link on the right will take you
directly to the Windows installer.
After you install PuTTY, open it up. You'll need to enter a hostname to
login to. The link on the right will take you to a list of student lab
hostnames\marginnote{You can find a list here: \LabHostnamesLink}
(such as
\hostname{batgirl.eecs.northwestern.edu}
or
\hostname{hush.eecs.northwestern.edu}). Ensure SSH is selected, then
press Open. You should get some sort of message asking whether or not
you trust the host. Press yes. From here, login as your EECS username
(probably the same as NetID), and your EECS password (not necessarily
your NetID password). You should now be logged into one of the
Northwestern EECS boxes!

Note that you can configure PuTTY so that you don't have to do all of
this every time.

If you would like to try to configure PuTTY to proxy jump so that
you don’t have to use the VPN, try \href{\PuTTYhelpUrl}{these
instructions}.

\subsection{Mac/Linux}

For those of you on Mac or Linux, everything you need is already
installed. Open up your terminal\marginnote{Mac users: search
for ``terminal'' in Spotlight} and at the prompt
type a single command of the form

\begin{CmdLine*}
  \C ssh \metavar{eecs-id}@\metavar{eecs-host}.eecs.northwestern.edu\\
\end{CmdLine*}
\marginnote[-\baselineskip]{Don't type the \prompt. That stands for your
shell prompt.}

\noindent where \metavar{eecs-id} is your EECS username (probably your NetID)
and \metavar{eecs-host} is replaced by one of the EECS hostnames from the list
of student lab hostnames%
\marginnote{\LabHostnamesLink}
(such as \hostname{alfred.eecs.northwestern.edu} or
\hostname{robin.eecs.northwestern.edu}).

You should get a message saying that the authenticity of the host
can't be established, and you should be asked if you want to continue
connecting.  Type ``yes'' as prompted and press Enter. Now type in your
EECS account password (not necessarily your NetID password), press
Enter again, and you should be logged in remotely!

You can configure ^ssh^ to proxy jump automatically, and to use a
\href{https://en.wikipedia.org/wiki/Key_authentication}{cryptographic
key for authentication} so that you don’t have to type your password. If
you want to try this, run the following command in your terminal and
follow the instructions:

\begin{CmdLine*}
  \C bash -c "\$(curl -fsLS \setupScriptLink)"\\
\end{CmdLine*}
\marginnote[-\baselineskip]{If you’d like to see the script, run just
  the \progname{curl} command that’s inside the parentheses (not including
the parentheses), or click the link.}

Notes: 1) Run this \emph{locally} on your own computer, not while logged
into a remote workstation. 2) Type or paste that command \emph{exactly}
as written.

\section{Basic shell navigation}

There are a few basic commands we will be using frequently throughout
this exercise in our shell: \progname{cd}, \progname{ls}, and
\progname{pwd}, and \progname{man}.

\progname{cd} stands for ``change directory,'' and is used to change the
current directory we are looking at in our shell (our working
directory). You can think of a directory as a folder from your regular
interactions with your computer. For example the command \cmdline{cd
Documents} \marginnote{As usual, don't type the \prompt.} will look for a
directory inside our current directory called \filename{Documents}, and
if it exists, our working directory will become that
\filename{Documents} directory. If you ever want to go back to your home
directory, the command \cmdline{cd} with no argument will switch your
working directory back to your home directory. The command \cmdline{cd
..} will switch your working directory up one level from where you
currently are.

\progname{pwd} stands for ``print working directory,'' and is used to
print out the current working directory of your shell. For example, if
you have been navigating around for a while and you are lost you can
type in the command \cmdline{pwd} and you will see your directory
printed out into the shell.

\progname{ls} is short for the word ``list,'' and is used to list the contents
and subdirectories within your current working directory. You can type
the command \cmdline{ls} into your shell, and you will see all files and
directories within your current working directory.

Play around with these three commands for a few minutes in your shell,
and see what directories and files already exist on your EECS box!

\progname{man} is short for ``manual,'' and is used to access the system
manuals. For example, you can read the manual pages for \progname{pwd}
and \progname{ls} by running the commands \cmdline{man pwd} and
\cmdline{man ls}.\marginnote{Hit \keycombo{q} to quit.} To learn about
man, you can of course run \cmdline{man man}.

Once you are done playing around, type \cmdline{cd} in to navigate back
to your home directory. We will be making a new directory for this lab
using the \progname{mkdir} command.

\section{Creating new files}

\progname{mkdir} stands for make directory, and is used to create a new
directory within our current working directory. For example,
\cmdline{mkdir fun-project} will create a new directory inside our
current one called \filename{fun-project} that we can \progname{cd} into
if we so desire. We can create hierarchies of directories to keep our
files well organized.

Create a new directory inside your home directory called
\filename{lab01-dir}. Change your current working directory to
\filename{lab01-dir}, and we will now practice editing and compiling
some C source files!

The \cmdline{emacs} command in the shell will open up the Emacs text%
\marginnote{Text editor preferences can be a fairly contentious issue
among software engineers, and if you already have experience with one
of Vim or Emacs, feel free to use whichever you already have
experience with instead of Emacs. However, for the purpose of this
class, we will be teaching using Emacs. Emacs can also seem scary at
first, but after you learn a few simple commands, it will quickly
start making sense. } editor. (On Mac/Linux, you will probably want to
use \cmdline{emacs} to avoid starting the X Window System..) Pass in
a file that you want to edit (even if it hasn't been created yet), and
you can start editing that file! For example type \cmdline{emacs
my\_code.c} and you can start editing a file called
\filename{my\_code.c} within your current working directory.

Inside your \filename{lab01-dir} directory create and open a file
using Emacs called \filename{animals.txt}. Note that the .c file
extension is what we will be using to indicate C files. You will see a
text editor pop up that does not look dissimilar to a
\progname{Notepad.exe} or \progname{TextEdit.app} editor from your
Windows or Mac. However, you will notice that clicking a location using
your cursor will not move your cursor to where you click :(

Inside this text editor, type in a list of your 3 favorite animals. Once
you have typed in your list, you are going to want to save your file so
you can use it later. On Emacs, saving is slightly different than other
programs. Instead of using \keycombo{Command-} or \keycombo{Ctrl-s}, you
are going to use \keycombo{Ctrl-x} followed by
\keycombo{Ctrl-s}.\marginnote{If you are curious about more Emacs
  commands, there is a nice basic list here:
\url{http://www.cs.cornell.edu/courses/cs312/2003sp/handouts/emacs.htm}.
You can also run an Emacs tutorial inside Emacs. Press \keycombo{C-h
t} -- that is, \keycombo{Ctrl-h} followed by \keycombo{t} (no
\keycombo{Ctrl}).}
(In Emacs, this is spelled ``\keycombo{C-x C-s}.'') This will save your
file to your current working directory. Now, we want to close our Emacs
window and get back to our Unix shell. In order to close our Emacs
window, we will type \keycombo{C-x C-c} (that is, \keycombo{Ctrl-x}
followed by \keycombo{Ctrl-c}).

We can ensure that our file was properly created by using the
\progname{cat} command in the shell. \progname{cat} is short for
``catenate,'' and prints out contents of a given file.  \cmdline{cat
\metavar{filename}} will print the contents of the file to the shell. If
you run \cmdline{cat animals.txt} you should see the file you just
created on your shell.

\section{Setting up the CS 211 development environment}

As you saw in class, we will be using the \progname{fish} shell in CS
211, because it’s friendlier and generally easier than other shells.
However, \progname{fish} isn’t available by default on the EECS
workstations, so getting access to it requires a bit of setup.

Additionally, editing programs will be much nicer if you have at least a
basic Emacs configuration.

Both of these things are configured in \emph{hidden}\marginnote{On
Unix-line operating systems, include Linux and Mac OS, files whose name
begin with a period (\texttt{.}) are omitted from directory listings by
default. Passing the \texttt{-a} flag to \progname{ls} causes it to show
these hidden files as well.} files in your home directory. The
directories where \progname{tcsh} (the default shell on these
workstations) looks for programs (called the ``path'') is configured in
\filename{.tcshrc}, and Emacs looks for its configuration in a file
called \filename{.emacs}.

We’ve provided a script to create these files for you. If you already
have these files, it will first back them up and tell you the names of
the backup files. Here is the command to run:

\begin{CmdLine*}
  \C \plaintilde jesse/setup211\\
\end{CmdLine*}

\marginnote[-\baselineskip]{Remember not to type the
\href{https://bit.ly/2yJyeD2}{prompt}.}

\noindent The changes will take effect the next time \progname{tcsh}
restarts. You can log out and back in, or run the command \cmdline{exec
tcsh} to reload \progname{tcsh}.

Once you’ve done so, you should be able to start \progname{fish}:

\begin{CmdLine*}
  \C fish \\
\end{CmdLine*}

You only need to do the above setup once, but you’ll
want to run \progname{fish} each time you log in.


\section{Getting the files}

We provide archives of starter code for both this lab and the homework
assignments (as well as code from the lectures) on the EECS login boxes
in a place where you can access them to make a copy. You'll want to use
these rather than starting from scratch because they include the build
system we’ll be using (more on this below) as well as configuration
options that you need.

So, you will find lab code in found in \verb^~cs211/lab^,%
\marginnote{On Unix, \verb^~^ in front of a user's name is the path to
their home directory, so this means the \filename{lab/} subdirectory of
belonging to a user named ^cs211^.} and the code for this lab in
particular may be found in \verb^~cs211/lab/lab01.tgz^. But what is a ^.tgz^
file and how can you use it?\marginnote{The ^.tgz^ file extension is
used for “\href{https://en.wikipedia.org/wiki/Gzip}{gzipped}
\href{https://en.wikipedia.org/wiki/Tar_(computing)}{tarball},” which is
like the Linux equivalent of a .ZIP file. The name ``tar'' stands for
``tape archive,'' because it was invented when computer systems still
stored backups on magentic tape.}

In order to extract the contents of an archive into your current working
directory, the command is:

\begin{CmdLine*}
  \C tar -xkvf \metavar{archive}\\
\end{CmdLine*}

\noindent
This week, \metavar{archive} should be \verb^~cs211/lab/lab01.tgz^.

The letters after the hyphen are \emph{flags}, which specify various
options to the program:

\begin{description}

  \item[k] tells ^tar^ to \textbf keep any existing files rather than
    overwrite them, which is the default. You might want to make a habit
    of this so that you never accidentally replace your completed
    homework with starter code, but you can leave the ^k^ out if you
    feel like living dangerously.

  \item[x] tells ^tar^ you want to e\textbf{xtract} files from the
    archive, as opposed to \textbf creating an archive or just \textbf
    telling you what's in it.

  \item[v] tells ^tar^ to be verbose, meaning it will print out the
    names of the files that it extracts; so if you don’t like seeing
    that part, you can leave it out.

  \item[f] tells ^tar^ to extract from the file whose name follows (as
    opposed to from some other place, like its \emph{standard input}).

\end{description}

Now, you have our new directory with the files you need, so change your
directory to \filename{lab01/} using the \progname{cd} command.
Now list its contents using \progname{ls}, and notice that there is a
\filename{Makefile} file, and a \filename{src/} directory. The
\filename{Makefile} file is a \progname{make} configuration file which
you won't have to worry about too much right now; the \filename{src/}
directory contains the file \progname{hello.c}, which we have provided
you.

\section{Using our build system}

As briefly mentioned in class, \progname{make} is our build system we
will be using for the first few weeks of the course at least.  We will
usually be giving some sort of starting structure for the projects you
will work on, and right now is no exception.

If you are the directory with the \filename{Makefile}\marginnote{This
means you should see it when you run \progname{ls}.}
then can build your program using the command \progname{make}.

% Linux provides a command called
% \progname{curl} that we will use to download a
% ``tarball'', and
% also a command \progname{tar} to extract the contents of the archive.
% \progname{curl} takes a URL as a command-line argument, downloads the
% file, and sends the result to its output; \progname{tar} reads an
% archive from its input and extracts it. We can use a Unix pipe, written
% with the \verb!|! character, to connect \progname{curl}’s output to
% \progname{tar}’s input:

% \begin{CmdLine*}
%   \C curl \metavar{url} | tar zxvk\\
% \end{CmdLine*}

% \noindent In this case, we will be getting our file from this url:
% \url{\ThisLabTgz}. However, the shell setup we did earlier defined
% several \emph{environment variables} containing URLs of various parts of
% the CS 211 website. Thus, our command will be:

% \begin{CmdLine*}
%   \C curl \$LAB211/\ThisLabBase.tgz | tar zxvk\\
% \end{CmdLine*}

The basic purpose of \progname{make} is to build your project into an
executable file. In your build directory, each time you update your
code, you can run

\begin{CmdLine*}
  \C make \metavar{target-name}\\
\end{CmdLine*}

\noindent to create your executable called \metavar{target-name}.  In
this case, run \cmdline{make hello} to build a program called
\filename{hello}, which will be put in the current directory. You can
run the program like this:

\begin{CmdLine*}
  \C ./hello\\
\end{CmdLine*}

\noindent
This should spit out a nice greeting.

\section{Updating our code}

So, we gave you a basic function and you were able to run it, but how do
you change the code?

Open up the \filename{src/hello.c} file using Emacs, and edit it so it
now says ``Aloha 211 student!'' instead of ``Hello 211 student!.'' Make
sure to save and exit Emacs.\marginnote{Remember \keycombo{C-x C-s} to
save and \keycombo{C-x C-c} to exit.}

Then try running \cmdline{./hello} again. Did anything change?

The reason why you still see ``Hello 211 student!'' on your screen is
because while you changed your C++ code, your computer doesn't
understand the C++ code, but only the machine code you create by using
\progname{make}.  So now, run \cmdline{make hello} once again, and
try \cmdline{./hello}. Notice how you now have the correct output!
Each time we want to change our code, we are going to need to remember
to rebuild our executable. Don't worry if you have error messages your
first few times trying to write new code, this is completely normal.
Even the best developers in the world usually need a few tries before
they can properly build their files, so just take a deep breath, and try
and figure out what went wrong.

\section{Conclusion}

Knowing how to use the shell is an extremely important tool in computer
science. Don't worry if it is still hard for you to use, like much of
life, it is one of those things you'll just need to practice with until
it seems much more familiar!  On your own time, it would be a good idea
to continue learning more about the shell and playing around with some
more commands.\marginnote{A good resource for some basic commands is
  here:
\url{http://www.computerworld.com/article/2598082/linux/linux-linux-command-line-cheat-sheet.html}.}
Of course, come to office hours or post on Campuswire with any questions or
if you want any more challenges!

\section{Useful links}

\begin{description}

\item[EECS login server hostnames]
\LabHostnamesLink

\item[Simple command line cheat sheet]
\url{http://www.computerworld.com/article/2598082/linux/linux-linux-command-line-cheat-sheet.html}

\item[PuTTY MSI installer]
\url{\PuTTYurl}

\item[PuTTY ``proxy jump'' instructions]
\url{\PuTTYhelpUrl}

\item[Nice Emacs guide]
\url{http://www.cs.cornell.edu/courses/cs312/2003sp/handouts/emacs.htm}

\item[Nice Vim guide]
\url{http://www.openvim.com/}

\item[VPN instructions]
\url{\GlobalProtectUrl}

\end{description}

\end{document}
