\documentclass{tufte-handout}

\usepackage{211lab}
\LabInfo{6}{Type Racer}

\begin{document}

\maketitle

Today we will be looking at another a C++ program using the GE211 game
engine in a slightly more advanced example game. The concept of the game
is simple enough: Words appear on screen as letters in circles, and you
to type the letters you see on the keyboard in order.

This game uses the model--view--controller pattern not-yet-described in
class, which allows defining the look, user interaction, and ``business
logic'' of an interactive program as separate components. Provided are
the \texttt{Model} class which defines the internal game state, the
\texttt{View} class for rendering game to the screen, and the
\texttt{Controller} class for reacting to user input and tieing it all
together.

\CxxPrelims*

\newpage

\section{General idea}

You have been given a fully functioning Type Racer that loads a
dictionary file (which can be found at
\filename{Resources/dictionary.dat})\marginpar{Your computer’s filesystem
may be \href{https://en.wikipedia.org/wiki/Case_sensitivity}{case
sensitive or insensitive}, and if case insensitive it may still be
\href{https://en.wikipedia.org/wiki/Case_preservation}{case preserving}.
If you aren’t careful, this can cause trouble if you share code with
someone else whose filesystem is different from yours. If you name your
resources directory \filename{resources/} instead of
\filename{Resources/} then it may work on some systems but fail to find
your resource files on others.}, and then displays each word from the
dictionary---in order---as letters inside colored circles. The player's
goal is to type the word, and the controller takes keyboard input to
update the player's progress through the word. The circles start out
yellow, and as the player progresses through the word, each circle
changes to green for a correctly typed letter, or red for a mistyped or
timed-out letter. Upon finishing a word the game loads the next word,
until all words in the dictionary have been exhausted. Try to identify
these components and trace their logic in the source code provided
before continuing.

\section{Randomize the dictionary}

In \filename{controller.cpp}, one of the constructors for the
\verb^Controller^ class calls a helper function \verb!load_dictionary()!
for loading the dictionary file into a \verb!std::vector<std::string>!.
Since the dictionary file is alphabetized and the model goes through the
word vector in order, this means that you see the same words, starting
with ``a,'' every time.

Your job is to modify the code of the \verb^Model^ and \verb^Controller^
classes to randomize the order of the words after the words are read in.
You will do this with a Fisher-Yates shuffle, which is a simple and
efficient algorithm for randomly permuting the order of a vector. The
algorithm\marginpar{
      In other words, if the vector has length $n$, first you choose a
      random element from index $0$ to $n - 1$ to put in position 0.
      Then choose a random element from index $1$ to $n - 1$ to put in
      position 1, and so on.
   } is:

{
\sffamily
\begin{tabbing}
\qquad\=\+
  \textbf{procedure} $\textsc{Shuffle}(v{:}\;\textsf{vector})$: \\
\qquad\=\+
  \textbf{for} $i$ \textbf{in} $0$
        \textbf{to} $\mathrm{len}(v) - 2$: \\
\qquad\=\+
  $r \gets \textrm{a random integer from between $i$ and $\mathrm{len}(v) - 1$}$; \\
  $\mathrm{swap}(v[i], v[r])$
\end{tabbing}
}

We don't want the model to randomize the words unconditionally, because
that would make testing too difficult. So instead, the shuffling itself
should happen in a new member function of the model at the controller's
request. Here's one way you can do it:

\begin{enumerate}

  \item Add a public member function to the \verb^Model^ class whose
    purpose is to shuffle the \verb^dictionary_^ vector. For a source of
    randomness, this function should take a reference to a
    \href{https://tov.github.io/ge211/classge211_1_1_random.html}
    {\texttt{ge211::Random}}. If \verb^rng^ is a \verb^ge211::Random&^
    then you can generate a random integer between \verb^a^ and \verb^b^
    (inclusive) with the call \verb^rng.between(a, b)^.

  \item Add a call to your shuffling function to the body of the
    \texttt{Controller::\-Controller(std::string const\&)} constructor.
    That way, when the game reads words from a file their order is
    randomized, but you can also avoid the shuffling by providing the
    vector directly.

\end{enumerate}

\section{Keep score}

Another thing that would be nice for this game is to add score keeping
of some kind. You could give $2$ points for every correct letter typed
and $10$ points for every word completed with no errors, but take away
$5$ points for every incorrect letter typed and $1$ point for every
letter timed out. When the game ends, have it stop and display the score
instead of looping on the word ``gameover.''

Here's a plan:

\begin{enumerate}
  \item Add a private\marginpar{
    Don't write a ``setter,'' because no one needs to set the score
    from outside the model.
    } member variable to the model to hold the score,
    and a public member function to allow the view to access it.

  \item Figure out how to detect the scoring events in the model code,
    and update\marginpar{
      Define constants for the event values. No magic numbers!
    } the score for each.

  \item Add private
    \href{https://tov.github.io/ge211/classge211_1_1_font.html}{\texttt{ge211::Font}}
    and
    \href{https://tov.github.io/ge211/classge211_1_1sprites_1_1_text__sprite.html}{\texttt{ge211::Text_sprite}}
    member variables to the view class. For creating the font, note that
    \texttt{"sans.ttf"} is included with GE211. For the text sprite, the
    initial text should be the number 0.

  \item Modify \texttt{View::draw} to:
    \begin{enumerate}
      \item
        \href{https://bit.ly/2SZr7N0}{reconfigure}
        the text sprite to contain the current score, and
      \item add the score sprite to the sprite set.
    \end{enumerate}
    In order\marginpar{
      You might also modify the view to keep track of the last score that it
      saw, so that it only needs to reconfigure the text sprite when the score
      changes.
    } to reconfigure the text sprite, you will need to create
    a
    \href{https://tov.github.io/ge211/classge211_1_1sprites_1_1_text__sprite_1_1_builder.html}{\texttt{ge211::Text_sprite::Builder}}
    and then add text to it. It looks something like this:
\begin{Code}
my_text_sprite_.reconfigure(
    ge211::Text_sprite::Builder(my_font_) << my_value)
\end{Code}
\end{enumerate}

\section{Testing}

It's hard testing whether your shuffle is producing permutations
uniformly. You could do repeated trials and check that you get a
reasonable distribution, but that's fairly tricky as soon as $n > 2$.
Easier, however, is to use
\href{https://en.cppreference.com/w/cpp/algorithm/is_permutation}
{\texttt{std::is_permutation()}} to check that the result of your shuffle
is a permutation of the original. \texttt{std::is_permutation()} can take
two begin--end pairs of \emph{iterators} to compare, and
\verb!std::vector<>! has public member functions \texttt{begin()} and
\texttt{end()} for getting such iterator pairs. Thus, you can check
whether two vectors \texttt{v} and \texttt{w} are permutations of each
other with:

\noindent
\bgroup\ttfamily
\strut~~~~CHECK(~std::is_permutation(v.begin(),~v.end(),
\marginpar{\rmfamily
  \texttt{std::is_permutation()} also has a three-argument form where
  you can omit the second end, provided you know the ranges are the same
  size. I guess you know that here, but this version seemed easier to
  explain.%
  } \\ \noindent
\strut~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~w.begin(),~w.end())~);
\egroup\par\medskip

You should also test that your model detects scoring events and keeps
score properly.

\section{Other ideas}

What if the game displayed a count-down timer for each letter's timeout?
(How can you format seconds with one decimal digit
\href{https://en.cppreference.com/w/cpp/io/manip/setprecision}{using
C++'s iotreams}?) Or instead of the numeric time, what if it showed a
bar whose length shrunk as the time ran out? (Does that require creating
a new \texttt{Rectangle_sprite} to change its size, or can you scale it
by passing a
\href{https://tov.github.io/ge211/classge211_1_1geometry_1_1_transform.html}
{\texttt{ge211::Transform}} to
\href{https://bit.ly/2SDxmHC}
{the four-argument form of \texttt{ge211::Sprite_set::add_sprite()}}?)

Can you make the game time the entire word for
\texttt{2 * current_word_.size()} seconds
instead of 2 seconds per letter as it does
now? (Then you'd really want some kind of display of the time
remaining.)

\end{document}
